A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-29 paper.tex
Improvements to sections 3-5

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine $T'$ with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined game tree}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
        \State $\hat{T} \gets $ \Call{extend}{$T$} \Comment{Extend the tree with unfixed actions}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$\hat{T}$}}{\Call{\textoverline{treeFormula}}{$\hat{T}$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_{\hat{T}}) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, \hat{T}$}}{\Call{\textoverline{learn}}{$\sigma, \hat{T}$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{nodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

@@

    \caption{Bounded Synthesis}
    \label{alg:bounded}
\end{algorithm}

@@

We instead present an approach that iteratively solves games of increasing
bound while learning states with interpolation over abstract games. We show
that reaching a fixed point in learned states is sufficient for completeness.

@@

\begin{algorithm}[h]
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets \top$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkBound}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkBound}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
    \end{algorithmic}
    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

@@

We modify the bounded synthesis algorithm to learn a subset of states from
failed attempts to find candidate strategies. When \textsc{findCandidate}
cannot find a strategy for an abstract game tree beginning at a state $s$, we
know that $s$ is a losing state. We can also attempt to learn additional losing
states from the tree via interpolation.

\begin{figure}
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$b$}
                };
%%%                node [above=4pt] {$\langle s', 1 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
\end{figure}

Consider the controller-losing game tree, $T$, in
Figure~\ref{fig:interpolatetree}. We choose any leaf node $l$ of $T$ and its
parent $n$. We then split the tree at $n$ such that both slices $T_1$ and $T_2$
contain $n$. Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$
except $l$, and $T_2$ (Figure~\ref{fig:treef2}) contains only $n$ and $l$ - a
single round of the game.  There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(T)$ is unsatisfiable, and by extension $s \land
\textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$ must also be
unsatisfiable. This enables the construction of an \emph{interpolant} that
captures losing states at node $n$.

Given two formulas $F_1$ and $F_2$ such that $F_1 \land F_2$ is unsatisfiable,
it is possible to construct a Craig interpolant~\cite{craig1957} $\mathcal{I}$
such that $F_1 \to \mathcal{I}$, $F_2 \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $F_1$ and $F_2$.
An interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $F_1 \land F_2$ \cite{pudlak1997}.

We can construct an interpolant with $F_1 = s \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$. The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By construction
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
must lose to the action in the edge between $n$ and $l$, the edge labelled by
action $b$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding the error state.

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{$B^M$}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

@@

When we discover losing states for the controller we record them in a set of
states that \emph{must} be bad, $B^M$. This set is initialised to the error
state at the beginning of the algorithm and grows as losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (see Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with
$B^M$. This enables further interpolants to be constructed by recursively
splitting more nodes from $T_1$ since the states that lost to $T_2$ are now
blocked by $B^M$.

It is more complicated to learn states from environment-losing abstract game
trees. States are only losing for the environment for a particular bound. For a
state learned at a node $n$ it is losing at $\textsc{bound}(n)$. Therefore we
maintain an array of state sets indexed by the bound at which that set of
states is losing for the environment. Instead of accumulating states, as we do
in $B^M$, we instead remove states from the array $B^m$. This array is
initialised with $\top$ and contains only states that \emph{may} lose for the
controller. The modified \textsc{\textoverline{treeFormula}} function
constrains the environment to may-losing states, which also allows for
recursive learning over the entire tree.

We compile interpolants into DNF cubes by constructing a temporary BDD. This
smaller representation is necessary for maintaining the size of formulas passed
to the SAT solver. However, it can occasionally lead to state explosion.

The recursive learning algorithms, \textsc{learn} (for the controller) and
\textsc{\textoverline{learn}} (for the environment), are detailed in
Algorithm~\ref{alg:learn}. When executing \textsc{findCandidate} in the
unbounded solver, lines 19 and 20 become active and call into the learning
procedures. The unbounded algorithm performs bounded synthesis for iteratively
increasing maximum bounds until the initial state is in $B^M$ (environment
wins) or $B^m$ reaches a fixed point (controller wins). We prove correctness in
the next section.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Require $\sigma(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $\sigma(X_T) \land B^M$
        \Function{learn}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$\sigma, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $\sigma(X_T) \land B^m[$\Call{bound}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{bound}{$n$}}
                \State $B^m[i] \gets B^m[i] \land \lnot \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$\sigma, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

@@

Line~(11) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has depth
1.  Consider formulas $F_1=\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ and $F_2 =
\textsc{\textoverline{treeFormula}}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $\sigma(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

Intuitively, the interpolant computed in line~(13) overapproximates the set of
states reachable from $\sigma$ by following the tree from the root node to $n$,
and underapproximates the set of states from which the environment loses
against tree $T_2$.

@@

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{bound}(n)]$ makes $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

At the second last recursive call to \textsc{\textoverline{learn}}, tree $T_1$
is empty, $n$ is the root node, $\textsc{\textoverline{treeFormula}}(T_1)
\equiv B^m[\textsc{bound}(T)](X^T)$; hence $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1) \equiv \sigma(X_T) \land
B^m[\textsc{bound}(T)](X^T) \equiv \bot$.  Thus the second postcondition of
\textsc{\textoverline{learn}} holds.

The proof of \textsc{learn} is similar to the above proof of \textsc{learn}. An
interpolant constructed from $F_1=\sigma(X_T) \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$ has the property $\II \land F_2 \equiv
\bot$ and the precondition ensures that the controller is unable to force the
game into $B^M$ playing against the counterexample moves in $T_2$. Thus adding
$\II$ to $B^M$ maintains the must-invariant satisfying the first postcondition.

Likewise, in the second last recursive call of \textsc{learn} with the empty
tree $T_1$ and root node $n$: $\textsc{treeFormula}(T_1) \equiv \lnot
B^M(X_T)$.  Hence $\sigma(X_T) \land \textsc{treeFormula}(T_1) \equiv
\sigma(X_T) \land \lnot B^M(X_T) \equiv \bot$. Therefore $\sigma(X_T) \land
B^M(X_T)$, the second postcondition, is true.

@@

The optimisation is relatively simple and is inspired by a common greedy
heuristic for minimising $\texttt{unsat}$ cores. $I$ is a value assignment to
each variable in $X$. If the environment does not win for $\langle I, k
\rangle$ then we attempt to solve for a generalised version of $I$ by removing
one assignment at a time. If the controller can win from the larger set of
states then we continue generalising without it. In this way we learn more
states by increasing the reachable set. In our benchmarks we have observed that
this optimisation is beneficial on the first few iterations of
\textsc{checkBound}.

>>>>

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine $T'$ with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined game tree}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
        \State $\hat{T} \gets $ \Call{extend}{$T$} \Comment{Extend the tree with unfixed actions}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$\hat{T}$}}{\Call{\textoverline{treeFormula}}{$\hat{T}$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_{\hat{T}}) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    %\State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$s, \hat{T}$}}{\Call{\textoverline{learn}}{$s, \hat{T}$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{nodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

@@

    \caption{Bounded synthesis}
    \label{alg:bounded}
\end{algorithm}

@@

We instead present an approach that iteratively solves games of increasing
bound while learning winning states from abstract games using 
interpolation. We show
that reaching a fixed point in learned states is sufficient for completeness.

@@

@@

We extend the bounded synthesis algorithm to learn states losing for one of the players from
failed attempts to find candidate strategies.  The learning procedure kicks in whenever
\textsc{findCandidate} cannot find a candidate strategy for an abstract game tree with initial
state $s$.  We know that $s$ is a losing state.  We can learn additional losing
states from the tree via interpolation.  This is achieved in lines~18--20 in Algorithm~\ref{alg:bounded},
enabled in the unbounded version of the algorithm, which invoke \textsc{learn} or
\textsc{\textoverline{learn}} to learn controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$b$}
                };
%%%                node [above=4pt] {$\langle s', 1 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
\end{figure}

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Require $s(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $s(X_T) \land B^M$
        \Function{learn}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$s, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $s(X_T) \land B^m[$\Call{bound}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{bound}{$n$}}
                \State $B^m[i] \gets B^m[i] \setminus \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$s, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

Consider the game tree $T$ in
Figure~\ref{fig:interpolatetree}, whose edges are labelled with environment
actions.  Assume that there does not exist a partial strategy for the controller
winning against this tree and that the \textsc{findCandidate} function could not
find a candidate strategy.
We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose children
are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree at $n$ such that both slices $T_1$ and $T_2$
contain a copy of $n$ (line~4). Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$
except $n$'s children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its children.  
There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(T)$ is unsatisfiable.  By construction, 
$\textsc{treeFormula}(T) = \textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$
and hence $s \land
\textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$ is also 
unsatisfiable. This enables the construction of an \emph{interpolant} that
captures losing states at node $n$.

Given two formulas $F_1$ and $F_2$ such that $F_1 \land F_2$ is unsatisfiable,
it is possible to construct a Craig interpolant~\cite{craig1957} $\mathcal{I}$
such that $F_1 \to \mathcal{I}$, $F_2 \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $F_1$ and $F_2$.
An interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $F_1 \land F_2$~\cite{pudlak1997}.

We can construct an interpolant with $F_1 = s \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$ (line~5). The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By construction
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
are losing against abstract game tree $T_2$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding error states.

We have discovered a set $\mathcal{I}$ of states losing for the controller.  
We record these newly discovered states in a set $B^M$ of
``bad'' states that \emph{must} be losing for the controller (Algorithm~\ref{alg:learn}, line~6). This set is initialised to the error
set $E$ and grows as new losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with
$B^M$. This enables further interpolants to be constructed by the learning procedure 
recursively splitting more nodes from $T_1$ (Algorithm~\ref{alg:learn}, line~7) since the states that lost to $T_2$ are now
blocked by $B^M$.

\begin{algorithm}[t]
    \caption{Amended tree formulas for Controller and Environment}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{$B^M$}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

@@

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialized to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{bound}(n)$ in line~13, Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

%It is more complicated to learn states from environment-losing abstract game
%trees. 
%States are only losing for the environment for a particular bound. For a
%state learned at a node $n$ it is losing at $\textsc{bound}(n)$. Therefore we
%maintain an array of state sets indexed by the bound at which that set of
%states is losing for the environment. Instead of accumulating states, as we do
%in $B^M$, we instead remove states from the array $B^m$. This array is
%initialised with $\top$ and contains only states that \emph{may} lose for the
%controller. 
The modified \textsc{\textoverline{treeFormula}} function (\ref{alg:unboundedTreeFormula})
constrains the environment to may-losing states, which also allows for
recursive learning over the entire tree.

%We compile interpolants into DNF cubes by constructing a temporary BDD. This
%smaller representation is necessary for maintaining the size of formulas passed
%to the SAT solver. However, it can occasionally lead to state explosion.

%The recursive learning algorithms, \textsc{learn} (for the controller) and
%\textsc{\textoverline{learn}} (for the environment), are detailed in
%Algorithm~\ref{alg:learn}. When executing \textsc{findCandidate} in the
%unbounded solver, lines 19 and 20 become active and call into the learning
%procedures.

Figure~\ref{alg:unbounded} shows the main loop of the unbounded synthesis algorithm.
The algorithm invokes the modified bounded synthesis procedure with increasing bound $k$
until the initial state is in $B^M$ (environment wins) or $B^m$ reaches a fixed point 
(controller wins). We prove correctness in the next section.

\begin{algorithm}[h]
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets \top$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkBound}{$k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $I \not\in B^m[k]$ if there exists a winning controller strategy with bound $k$
        \Ensure $I \in B^M$ if there exists a winning environment strategy with bound $k$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkBound}{$k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, I, \emptyset$}
        \EndFunction
    \end{algorithmic}
    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

@@

@@

Line~(11) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has depth
1.  Consider formulas $F_1=s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ and $F_2 =
\textsc{\textoverline{treeFormula}}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $s(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

Intuitively, the interpolant computed in line~(13) overapproximates the set of
states reachable from $s$ by following the tree from the root node to $n$,
and underapproximates the set of states from which the environment loses
against tree $T_2$.

@@

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{bound}(n)]$ makes $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

At the second last recursive call to \textsc{\textoverline{learn}}, tree $T_1$
is empty, $n$ is the root node, $\textsc{\textoverline{treeFormula}}(T_1)
\equiv B^m[\textsc{bound}(T)](X^T)$; hence $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1) \equiv s(X_T) \land
B^m[\textsc{bound}(T)](X^T) \equiv \bot$.  Thus the second postcondition of
\textsc{\textoverline{learn}} holds.

The proof of \textsc{learn} is similar to the above proof of \textsc{learn}. An
interpolant constructed from $F_1=s(X_T) \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$ has the property $\II \land F_2 \equiv
\bot$ and the precondition ensures that the controller is unable to force the
game into $B^M$ playing against the counterexample moves in $T_2$. Thus adding
$\II$ to $B^M$ maintains the must-invariant satisfying the first postcondition.

Likewise, in the second last recursive call of \textsc{learn} with the empty
tree $T_1$ and root node $n$: $\textsc{treeFormula}(T_1) \equiv \lnot
B^M(X_T)$.  Hence $s(X_T) \land \textsc{treeFormula}(T_1) \equiv
s(X_T) \land \lnot B^M(X_T) \equiv \bot$. Therefore $s(X_T) \land
B^M(X_T)$, the second postcondition, is true.

@@

The optimisation is relatively simple and is inspired by a common greedy
heuristic for minimising $\texttt{unsat}$ cores. Initial state $I$ assigns a value to
each variable in $X$. If the environment loses $\langle I, k
\rangle$ then we attempt to solve for a generalised version of $I$ by removing
one variable assignment at a time. If the environment loses from the larger set of
states then we continue generalising. In this way we learn more
states by increasing the reachable set. In our benchmarks we have observed that
this optimisation is beneficial on the first few iterations of
\textsc{checkBound}.
